Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2019-08-02 peakcharacterization_manuscript.Rmd
updated wording

$$frequency = intercept + slope* \frac{1}{\sqrt{mz}} $$
A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans. A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. Integer values (for use with the IRanges R package) are generated by multiplying each frequency value by 400 and rounding.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data, outside of meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference. The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb. Our instrument seems to have a mode of 0.5, as shown in Figure Xc. Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure Xd. This model seems to fit the known relationship between frequency and M/Z. This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points. M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z. Frequency based points suffer none of these drawbacks, and it turns out that the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

```{r mz_frequency_figure}
# Figure showing relationship of M/Z to frequency
# a - mean m/z vs diff m/z
# b - mean m/z vs diff frequency
# c - histogram of diff frequency
# d - mean m/z vs frequency

These models are calculated for each scan, and the intercepts from the models compared and examined for outliers. Outlier intercepts are determined from the interquartile range, and outliers removed. When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure 2a-c. Therefore, a single model based on the median intercept and slope is used for converting *all* remaining scan level data.

>>>>

$$frequency = intercept + x* \frac{1}{\sqrt{mz}} + y * \frac{1}{\sqrt[3]{mz}}$$
From the physical properties of the orbitrap, only the square root should be necessary, however, in practice we have found the combination of square and cube-roots to provide a better fit, likely due to issues with slight imperfections in the orbitrap geometry and control of the magnetic fields. A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans. A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. Integer values (for use with the IRanges R package) are generated by multiplying each frequency value by 400 and rounding.

@@

FT-MS data from the Thermo-Fisher Orbitrap instrument used to acquire the data does not provide any information about the raw frequency data. Outside of the meta-data, it merely contains the M/Z and intensity values for profile spectra acquired across multiple scans. However, it turns out that the frequency can be calculated by dividing the average M/Z of two adjacent points with their difference (Figure XA, red points reresenting the average M/Z of two adjacent points, red lines representing the difference between the two adjacent points). The differences in frequency should be relatively constant with respect to M/Z, in contrast to the differences of adjacent M/Z points, shown in Figure Xa and Xb. Our instrument seems to have a mode of 0.5, as shown in Figure XC and XD. Restricting to those points that fall into a narrow range of frequency differences (0.49 - 0.51), a model of frequency to M/Z can be generated (see Methods), an example of which is shown in Figure XE. This model seems to fit the known relationship between frequency and M/Z. This is useful, because some of the subsequent steps in our workflow use sliding windows where it is assumed that the sliding windows contain the same number of data points. M/Z point-to-point differences are not constant, but can be approximated by a lasso linear model. However, it is very difficult to create a lasso model with an intercept of 0, and then vary the width of sliding windows according to the M/Z difference at a particular M/Z. Frequency based points suffer none of these drawbacks, and the conversion from M/Z can be derived from the raw profile level data itself, which is incredibly useful.

```{r mz_frequency_figure}
drake::loadd(frequency_conversion)
frequency_conversion
```

These models are calculated for each scan, and the square root term from all scan level models checked for outliers based on the interquartile ranges across all scans in a sample. When scan specific models are used for conversion of M/Z to frequency, the peak order observed in M/Z space is not maintained, as shown in Figure XA. Therefore, a single model for all scans based on the median of each term in the model across scans is used for converting *all* remaining scan level data.
